Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-05-16 content.tex
Use 'workflow.png' instead of keeping 'workflow[version].png', git can control versions
Processed Rampal's feedback, fix #46

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{workflow3.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the generative tree prior 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed by a clock and a site model. 
    The user supplies one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a tuple of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected 
    to be used in the next step.
    The inference models (3) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the tree prior $\mathit{p_{I}}$ of the selected inference model (3).
    Analogously, a twin alignment is simulated from the twin phylogeny along with clock model $\mathit{c_{I}}$ and site model $\mathit{s_{I}}$ imported from the selected inference model.
    The twin pipeline mimics the procedure of the main pipeline, resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
We define a phylogeny $\tau$ is the combination of 
branching times $\Vec{t}$ and topology $\psi$.
Assuming that a non-standard generative prior $\mathit{p_{G}}$
produces a tree $\tau_{\mathit{G}}$,
with branching times $\Vec{t}_{\mathit{G}}$ and 
topology $\psi_{\mathit{G}}$, the twinning process T, in accordance with a chosen inference 
model $\mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving the 
topology $\psi_{\mathit{G}}$:

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{ | l | l | l | }
    \hline
    \textbf{Name} & \textbf{Description} & \textbf{Listing} \\
    \hline
    \verb;pir_run; & Run \verb;pirouette; & \ref{lst:pir_run} \\
    \verb;pir_plot; & Show the \verb;pirouette; results as a plot & \ref{lst:pir_plot} \\
    \verb;create_pir_params; & Create the \verb;pirouette; parameters & \ref{lst:create_pir_params} \\
    \hline
    \verb;create_alignment_params; & Create the alignment parameters & \ref{lst:create_alignment_params} \\
    \verb;create_twinning_params; & Create the twinning parameters & \ref{lst:create_twinning_params} \\
    \verb;create_experiment; & Create one experiment & \ref{lst:create_gen_experiment_explicit} \\
    \verb;create_error_measure_params; & Create the error measurement parameters & \ref{lst:create_error_measure_params} \\
    \hline
  \end{tabular}
  \caption{
    \texttt{pirouette}'s main functions, description and the number of the 
    listing in which it is used. 
  }
  \label{tab:functions}
\end{table}
\giovanni{I think the reference to the listings is a great idea. I don't know what to add more to this table. Probably (not 100\% sure) we should make all tables fit within the page limits.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error for both generative and best candidate inference models.
  }
  \label{fig:example_6}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
  }
  \label{fig:example_5}
\end{figure}
\giovanni{I find the figure to be too confused. What if we use instead different plots, and in each of them we just show one inference model's results plus its twin counterpart?}

@@

However, if the same procedure were repeated and performed on a distribution with a sufficient number of generative trees, it would constitute a quantitative and effective assessment of the quality of the inference.

A twin tree does not always result in a lower error distribution,
due to the stochasticity in generating the twin tree.
For example, of the second tree used in the 'Usage' section,
94\% of all twin trees resulted in an error distribution with a lower median.
\giovanni{If we write this some reviewer could argue: where do you see this? Do you have a plot to show it? Why don't you present such data about a bigger distribution, at least in an appendix?}

% Add: the effect of the alignment on this
% https://github.com/richelbilderbeek/pirouette_article/issues/45

\iffalse
\giovanni{We can probably just remove this.}
To give an idea of the insights \verb;pirouette; can give,
we pretend the final result was to be obtained from thousands of replicates.
We ...
\fi

>>>>

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the generative tree prior 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed by a clock and a site model. 
    The user supplies one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a tuple of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected 
    to be used in the next step.
    The inference models (3) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the tree prior $\mathit{p_{I}}$ of the selected inference model (3).
    Analogously, a twin alignment is simulated from the twin phylogeny along with clock model $\mathit{c_{I}}$ and site model $\mathit{s_{I}}$ imported from the selected inference model.
    The twin pipeline mimics the procedure of the main pipeline, resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b)
and 'twin alignment'
that will be analyzed in the same way as the true tree (and alignment).
We define a phylogeny $\tau$ is the combination of
branching times $\Vec{t}$ and topology $\psi$.
Assuming that a non-standard generative prior $\mathit{p_{G}}$
produces a tree $\tau_{\mathit{G}}$,
with branching times $\Vec{t}_{\mathit{G}}$ and 
topology $\psi_{\mathit{G}}$, the twinning process T, in accordance with a chosen inference 
model $\mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving the 
topology $\psi_{\mathit{G}}$:

@@

From the twin tree, we simulate an alignment following the same
history as the twin tree, with the addition that 
the same total number of mutations have
taken place from root sequence to the sequences in the present.
This addition assures there is the same amount of historical information
stored in the alignments of both the true and twin alignment.
We achieve this by simply simulating twin alignments up until we
obtain one that has the desired number of mutations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{ | l | l | l | }
    \hline
    \textbf{Name} & \textbf{Description} & \textbf{Listing} \\
    \hline
    \verb;pir_run; & Run \verb;pirouette; & \ref{lst:pir_run} \\
    \verb;pir_plot; & Show the \verb;pirouette; results as a plot & \ref{lst:pir_plot} \\
    \verb;create_pir_params; & Create the \verb;pirouette; parameters & \ref{lst:create_pir_params} \\
    \hline
    \verb;create_alignment_params; & Create the alignment parameters & \ref{lst:create_alignment_params} \\
    \verb;create_twinning_params; & Create the twinning parameters & \ref{lst:create_twinning_params} \\
    \verb;create_experiment; & Create one experiment & \ref{lst:create_gen_experiment_explicit} \\
    \verb;create_error_measure_params; & Create the error measurement parameters & \ref{lst:create_error_measure_params} \\
    \hline
  \end{tabular}
  \caption{
    \texttt{pirouette}'s main functions, description and the number of the 
    listing in which it is used. 
  }
  \label{tab:functions}
\end{table}
\giovanni{Probably (not 100\% sure) we should make all tables fit within the page limits.}
\richel{Nope, the publisher will do this for us in the end}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error for both generative and best candidate inference models.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_6}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_5}
\end{figure}
\giovanni{
  I find the figure to be too confused. 
  What if we use instead different plots, 
  and in each of them we just show one inference model's results plus its twin counterpart?
}
\richel{Will do}

@@

However, if the same procedure were repeated and performed on a distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the quality of the inference.
Also a twin tree does not always result in a lower error distribution,
due to the stochasticity in generating the twin tree.
